### 2020-02-06 UbiCompSubmission/bp_conclusion.tex
Extra revissions

%!TEX root = bioPrediction_main.tex
\section{Conclusions} 
Stress, awakeness and focus at work are highly relevant aspects when
it comes to productivity and well-being at the workplace. In this
paper, we presented the results of a study with 14 professional
knowledge workers in their workplace over an eight week period to
better understand how workers experience stress over time and to
examine the ability of biometrics to predict these
productivity-related aspects. The longitudinal and in situ placement of
the study support and extend previous work. Based on
twice collected daily survey responses, we observed that
although participants sometimes saw periods of sustained stress,
they would always return to a baseline stress level for them
at some point. We also observed that stress levels seldom spiked,
but when they did rise, the rise in stress tended to last more
than a day. In addition to the survey responses, we 
continually collected
biometrics data with which we were able to create a model that was able to
predict awakeness, stress, and focus.

>>>>

%!TEX root = bioPrediction_main.tex
\section{Conclusions} 
Stress, awakeness and focus at work are highly relevant aspects when
it comes to productivity and well-being at the workplace. In this
paper, we presented the results of a study with 14 professional
knowledge workers in their workplace over an eight week period to
better understand how workers experience stress over time and to
examine the ability of biometrics to predict these
productivity-related aspects. The longitudinal and in-situ placement of
the study support and extend previous work. Based on
twice collected daily survey responses, we observed that
although participants sometimes saw periods of sustained stress,
they would always return to a baseline stress level for them
at some point. We also observed that stress levels seldom spiked,
but when they did rise, the rise in stress tended to last more
than a day. In addition to the survey responses, we 
continually collected
biometrics data with which we were able to create a model that is able to
predict user stress, sleepyness and lack of focus with an up to 84.65\% improvement over the baseline.

### 2020-02-06 UbiCompSubmission/bp_intro_v4.tex
Extra revissions

Given the importance of understanding stress and its relationship and
effect on work, a number of studies have been conducted using a
variety of methods. Some have studied stress using invasive
techniques, such as measuring cortisol as a stress biomarker~\cite{piazza10} which are best suited
for laboratory environments.  Further research has shown the benefit of using
autonomic nervous system (ANS) activity by analyzing biometric signals
of the human body, such as blood pressure~\cite{kataoka00}.
These less invasive approaches can be used in longer-term field
studies, opening up the question of what stress
looks like in the wild.

@@

Our work builds upon previous work and extends it by performing an
eight-week study in the workplace with 14 knowledge 
workers. To the best of our knowledge, this is the 
longest in-situ study performed in a real-world 
environment of knowledge workers to analyze an extensive collection of biometric signals with experience samples examining real-time prediction of stress in the workplace~\cite{sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent,Parnin2011,Nakagawa2014,Radevski2015}.
Previous studies are predominantly controlled lab studies which used only a subset of the relevant biometric signals we captured, have been much more limited in duration, or are not focused on the workplace.

We collected biometric data as well as
experience samples on stress, and two related human aspects: focus and
awakeness. We collected participant reports on focus because the
challenges that contribute to stress levels can also make it hard for
knowledge workers to stay focused and the level of focus
can affect
productivity~\cite{mark2014bored}.  We collected participant reports
on awakeness because a lack of awakeness (sleepiness) can result in
undesired consequences on work~\cite{connor02}.  The participants in
our study, who perform various job functions for a research and
development group within a single large corporation, wore a novel, state-of-the-art
biometric armband sensor\footnote{Biovotion Everion
  sensor~\cite{everion}} with low invasiveness that captured heart-,
respiration- and skin-related measures. This modality was chosen to
ease longitudinal deployment in the field. Using machine learning, we
created classifiers and analyzed their ability to predict stress, focus, and awakeness levels. Considering all of these three human aspects helps us to consider whether one aspect might be easier to detect than another. If one aspect is easier to detect, it might serve
as a proxy or an indicator of the presence (or absence) of other aspects.

In our analysis, we identify several trends in day-to-day stress levels
that emerged over the course of our eight-week study.  The results of
our analysis also show that biometric signals collected from a single
minimally invasive sensor can be used to predict stress and its
related aspects accurately, with the abstract concept of focus
(predictably) being the hardest to detect. Our results further show
that knowledge workers' self-reported levels of stress, focus and
awakeness and their physiological manifestation and prediction can
vary substantially between individuals.
%Our results further show that overall combining biometric features with computer interaction information 
%results in an improvement
%over each of the training sets individually.
The main contributions of our work are:
\begin{itemize}
	\item A qualitative examination of how knowledge worker stress presents and fluctuates in the wild.
	\item The creation and analysis of measures for the automatic monitoring of knowledge workers' stress, focus and awakeneess in the workplace based on an eight-week field study with 14 office workers.
	%\item A thorough analysis of ways to increase performance, including the training sample size, time window to collect data, and the inclusion of computer interaction features.
	\item A discussion on the impact of applying this research to improve
the interaction of knowledge workers with their digital environment leading to an improvement in their productivity and well-being, besides a reflection on aspects that can be further improved in future studies.
    \end{itemize}

>>>>

Given the importance of understanding stress and its relationship and
effect on work, a number of studies have been conducted using a
variety of methods. Minimally invasive studies have shown the benefit of using
autonomic nervous system (ANS) activity by analyzing biometric signals
of the human body, such as blood pressure~\cite{kataoka00}.
These approaches can be used in long-term field
studies, opening up the question of what stress
looks like in the wild. 
In contrast, other studies have studied stress using invasive
techniques, such as measuring cortisol as a stress biomarker~\cite{piazza10}. These studies are best suited
for laboratory environments.

@@

Our work builds upon previous work and extends it by performing an
eight-week study in the workplace with 14 knowledge 
workers. To the best of our knowledge, this is the 
longest in-situ study performed in a real-world 
environment of knowledge workers to analyze an extensive collection of biometric signals with experience samples examining real-time prediction of stress in the workplace~\cite{sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent,Parnin2011,Nakagawa2014,Radevski2015}.
Previous studies are predominantly controlled lab studies which used only a subset of the relevant biometric signals we captured, have been much more limited in duration, or are not focused on a real-world environment0.

We collected biometric data as well as
experience samples on stress, and two related human aspects: focus and
awakeness. We collected participant reports on focus because the
challenges that contribute to stress levels can also make it hard for
knowledge workers to stay focused and the level of focus
can affect
productivity~\cite{mark2014bored}.  We collected participant reports
on awakeness because a lack of awakeness (sleepiness) can result in
undesired consequences on work~\cite{connor02}.  The participants in
our study perform various job functions for a research and
development group within a single large corporation. For this study they wore a novel, state-of-the-art
biometric armband sensor\footnote{Biovotion Everion
  sensor~\cite{everion}} with low invasiveness that captured heart-,
respiration- and skin-related measures. This modality was chosen to
ease longitudinal deployment in the field. Using machine learning, we
created classifiers and analyzed their ability to predict stress, focus, and awakeness levels. Considering all of these three human aspects helps us to consider whether one aspect might be easier to detect than another. If one aspect is easier to detect, it might serve
as a proxy or an indicator of the presence (or absence) of other aspects.

In our analysis, we identify several trends in day-to-day stress levels
that emerged over the course of our eight-week study.  The results of
our analysis also show that biometric signals collected from a single
minimally invasive sensor can be used to predict stress and its
related aspects accurately, with the abstract concept of focus
(predictably) being the hardest to detect. Our results further show
that knowledge workers' self-reported levels of stress, focus and
awakeness and their physiological manifestation and prediction can
vary substantially between individuals.
%Our results further show that overall combining biometric features with computer interaction information 
%results in an improvement
%over each of the training sets individually.
The main contributions of our work are:
\begin{itemize}
	\item A qualitative examination of how knowledge worker stress presents and fluctuates in the wild based on an eight-week field study with 14 office workers.
	\item The creation and analysis of measures for the automatic monitoring of knowledge workers' stress, focus and awakeneess in the workplace based on a machine learning model trained on biometric data and experience samples.
	%\item A thorough analysis of ways to increase performance, including the training sample size, time window to collect data, and the inclusion of computer interaction features.
	\item A discussion on the impact of applying this research to improve
the interaction of knowledge workers with their digital environment leading to an improvement in their productivity and well-being, besides a reflection on aspects that can be further improved in future studies.
    \end{itemize}

### 2020-02-06 UbiCompSubmission/bp_results_1.tex
Extra revissions

To investigate whether stress, focus and awakeness can be predicted in
the moment based on biometric measures.
we investigated classifiers trained
for each individual and across all participants. We report on the effectiveness of
these classifiers and the features that are important in prediction of stress, focus
and awakeness.

@@

From the start time of each survey response, we look back the amount of time 
that corresponds to each time window, and we create features for all of the 
biometric data available in that time window. For example, if a participant 
started a survey response at 11:05am, for the 30min time window, we create 
features using all of the available biometric data from 10:35am to 11:05am. 
For each time window, we calculate 10 statistical measurements from the 
biometric data to create 10 distinct features. Specifically, the 10 
statistical measurements are: mean, standard deviation, variance, median, 
percentile25, percentile75, interquartile range, maximum, minimum, and 
range. Thus, for each survey response, we generate a large number of 
corresponding features based on three factors: biometric measurement, time 
window, and statistical measurement. In addition to these biometric 
features, we also considered the time of day in which the questions were 
asked. These features are created to predict the responses described by the 
ground truth.

@@

\subsection{Individual Classifiers}
%\noindent\textit{Results of Individual Classifiers}\\
Since peoples' experience of stress, focus, and awakeness (as well as
their physiological manifestations) can vary substantially
(e.g.,~\cite{Hernandez11}), we first trained individual classifiers
for each participant (rather than a general one for all
participants). The results of our analysis are reported in
Table~\ref{tab:accuracy}. For our analysis, we report values of
accuracy, one of the most commonly used metric to compare performance,
as well as precision and recall of the classes of interest:
'stressed', 'not focused', and 'not awake'. Since the imbalance in the
data can lead to high accuracy values if a classifier always just
predicts the most likely/frequent class while ignoring the class of
higher importance and interest, precision and recall of the class of
interest are also important to
consider~\cite{yap2014,bhattacharyya_data_2011,Hernandez11}.
For some users (i.e., S1. as seen In Table \ref{responseDistribution}), the imbalance in their data was so extreme that even after adjusting by oversampling it was impossible to create a reasonable classifier. These scenarios are difficult to predict as any classifier will not have enough variance in its training data for the 'stressed' situation to adequately distinguish it from the non-stressed case.

>>>>

To investigate whether stress, focus and awakeness can be predicted in
the moment based on biometric measures,
we investigated classifiers trained
for each individual and across all participants. We report on the effectiveness of
these classifiers and the features that are important in prediction of stress, focus
and awakeness.

@@

From the start time of each survey response, we look back the amount of time 
that corresponds to each time window, and we create features for all of the 
biometric data available in that time window. For example, if a participant 
started a survey response at 11:05am, for the 30min time window, we create 
features using all of the available biometric data from 10:35am to 11:05am. 
For each time window, we calculate 10 statistical measurements from the 
biometric data to create 10 distinct features. Specifically, the 10 
statistical measurements are: mean, standard deviation, variance, median, 
25\textsuperscript{th} percentile, 75\textsuperscript{th} percentile, interquartile range, maximum, minimum, and 
range. Thus, for each survey response, we generate a large number of 
corresponding features based on three factors: biometric measurement, time 
window, and statistical measurement. In addition to these biometric 
features, we also considered the time of day in which the questions were 
asked. These features are created to predict the responses described by the 
ground truth.

@@

\subsection{Individual Classifiers}
%\noindent\textit{Results of Individual Classifiers}\\
Since peoples' experience of stress, focus, and awakeness (as well as
their physiological manifestations) can vary substantially
(e.g.,~\cite{Hernandez11}), we first trained individual classifiers
for each participant (as opposed to a general one for all
participants). The results of our analysis are reported in
Table~\ref{tab:accuracy}. For our analysis, we report values of
accuracy, one of the most commonly used metric to compare performance,
as well as precision and recall of the classes of interest:
'stressed', 'not focused', and 'not awake'. Since the imbalance in the
data can lead to high accuracy values if a classifier always just
predicts the most likely/frequent class while ignoring the class of
higher importance and interest, precision and recall of the class of
interest are also important to
consider~\cite{yap2014,bhattacharyya_data_2011,Hernandez11}.
For some users (i.e., S1. as seen in Table \ref{responseDistribution}), the imbalance in their data was so extreme that even after adjusting by oversampling it was impossible to create a reasonable classifier. These scenarios are difficult to predict as any classifier will not have enough variance in its training data for the 'stressed' situation to adequately distinguish it from the non-stressed case.

### 2020-02-06 UbiCompSubmission/bp_studyMethod.tex
Extra revissions

\subsection{Participants}
We recruited group of 14 professionals via personal contacts from a large power and automation company. The group is very diverse in terms of age, work experience, gender, and work responsibilities. All participants work primarily in an office environment, though half of the participants spend at least 10\% (and up to 50\%) of their time in a laboratory environment. Office workers are a population that generalizes to a variety of contexts, and including part-time laboratory workers guarantees that our participants have varying work patterns that include different levels of computer usage, as well as different levels of activity in both individual and collaborative tasks.

Of the 14 participants, 11 are male and 3 are female. The average participant age is ~40, with 5 in the age range 25-34, 7 in the age range 35-44, and 2 in the age range 55+. The participants have an average number of years of professional experience of ~12, with 2 having less than 5 years, 10 having 5-15 years, and 2 having more than 25 years. All participants work for a research organization within the company, but their job functions span line management, laboratory science, scientific research, technology evaluation, and software development.

\subsection{Procedure}
We performed an in-situ study over the course of eight weeks. For this, we first informed participants about the study purpose and procedure, handed out biometric sensors and introduced and explained the self-reporting to the participants. For 10 of the participants, we further installed a computer activity tracker on their company-issued laptops (note: four participants from the original sample declined for privacy reasons). After the initial setup, participants were asked to fill out 3 surveys each day for the following eight weeks (see Section~\ref{sec:Surveys}). At the end of the eight weeks, we collected the biometric sensors and performed a short follow-up interview on the study and the participants' experiences.

In the second month of the study, we offered participants two one-hour Tai Chi classes per week (each in the middle of the day) in addition to the daily experience sampling and asked them to attend one class a week. We chose to offer Tai Chi class for two reasons: first, to incentivise participants for their continued participation and for keeping them engaged; and second, for their said benefits of reducing stress based on our consultation with researchers in psychiatry, psychology and more specifically on mindfulness practices.

\begin{comment}
\subsection{Procedure}
\todo{This section does not work here. It is not really the study procedure, but rather the project procedure. Perhaps it can form the basis of an overview section after Field Study but before Analysis and Results.}

@@

We use state-of-the-art sensors to capture a stream of biometric data from each individual and a computer interaction tracker to collect computer interaction information during the time
of the study.

The participants were asked to fill a survey 3 times a day: two during the working day, and
one at the end of the day, where they rated their level of stress, focus, and awakeness during
the day; and their level of stress, awakeness, productiveness, and overall feeling at the end of the day.
This was conducted over an eight week period, which is
400\% longer than previous studies~\cite{zuger18,Muller16}.

We then pick a set of time windows (10sec, 20sec, 30sec, 45sec, 1min, 2min, 3min, 5min, 7.5min, 10min, 20min, 30min, 45min, 1hour, 2hour, 3hour) to analyze. We applied different statistical measurements (mean, standard deviation, variance, median, percentile25, percentile75, interquartile range, maximum, minimum, range) to each of the collected biometric signals and the computer interaction data within the time windows selected. We then 
compared the performance of four different machine learning 
algorithms to be able to predict the survey responses based in the biometric and 
computer interaction data.

Based in the initial corpus of training data, we create a machine learning model that is able to accurately predict the responses for stress, focus, and awakeness for each of the individuals wearing the sensors.
We are therefore able to infer the answer within the spectrum to each of these indicators without the need to ask the participants to provide an answer.
\end{comment}

@@

\subsubsection{Computer Interaction Data}
To gain a better understanding of our participants day-to-day work activities, we installed an open source computer interaction monitor (reference omitted for doubleblind). The monitor ran in the background on participants' computers and tracked the active windows, as well as the keyboard and mouse activity. Four of our participants opted out of this part of the study for privacy reasons (S6, S10, S12, and S13). Therefore, we only included the computer interaction data in our analysis for our more coarse-grained explanatory model but not in our analysis on the predictive model for stress in the moment.

\subsubsection{Surveys}
\label{sec:Surveys}
Following guidelines from previous studies~\cite{Lalle16,Panwar18,Luo18} and 
following the preferrences of extensive user piloting, we sent via 
text message a survey request to each participant two times per 
work day. Pilot participants preferred these over other means, in part, due 
to them being accessible and noticeable anywhere in the office.

>>>>

\subsection{Participants}
We recruited group of 14 professionals via personal contacts from a large power and automation company. The group is diverse in terms of age, work experience, gender, and work responsibilities. All participants work primarily in an office environment, though half of the participants spend at least 10\% (and up to 50\%) of their time in a laboratory environment. Office workers are a population that generalizes to a variety of contexts, and including part-time laboratory workers guarantees that our participants have varying work patterns that include different levels of computer usage, as well as different levels of activity in both individual and collaborative tasks.

From the 14 participants, 11 are male and 3 are female. The average participant age is ~40, with 5 in the age range 25-34, 7 in the age range 35-44, and 2 in the age range 55+. The average number of years of professional experience of the participants is ~12, with 2 having less than 5 years, 10 having 5-15 years, and 2 having more than 25 years. All participants work for a research organization within the company, but their job functions span line management, laboratory science, scientific research, technology evaluation, and software development.

\subsection{Procedure}
We performed a study over the course of eight weeks. This study includes the collection of the participant's biometric data, computer interaction data, and self-reporting surveys. We informed participants about the study purpose and procedure, handed out biometric sensors, and introduced and explained the self-reporting to the participants. For 10 of the participants, we further installed a computer activity tracker on their company-issued laptops (note: four participants from the original sample declined for privacy reasons).

After the initial setup, participants were asked to fill out 3 surveys each day for the following eight weeks (see Section~\ref{sec:Surveys}). At the end of the eight weeks, we collected the biometric sensors and performed a short follow-up interview on the study and the participants' experiences.

In the second month of the study, we also offered participants two one-hour Tai Chi classes per week (each in the middle of the day) and asked them to attend one class a week. We chose to offer Tai Chi class for two reasons: first, to incentivise participants for their continued participation and for keeping them engaged; and second, for their said benefits of reducing stress based on our consultation with researchers in psychiatry, psychology and more specifically on mindfulness practices.

\begin{comment}
%\subsection{Procedure}
%\todo{This section does not work here. It is not really the study procedure, but rather the project procedure. Perhaps it can form the basis of an overview section after Field Study but before Analysis and Results.}

@@

%We use state-of-the-art sensors to capture a stream of biometric data from each individual and a computer interaction tracker to collect computer interaction information during the time
%of the study.

%The participants were asked to fill a survey 3 times a day: two during the working day, and
%one at the end of the day, where they rated their level of stress, focus, and awakeness during
%the day; and their level of stress, awakeness, productiveness, and overall feeling at the end of the day.
%This was conducted over an eight week period, which is
%400\% longer than previous studies~\cite{zuger18,Muller16}.

%We then pick a set of time windows (10sec, 20sec, 30sec, 45sec, 1min, 2min, 3min, 5min, 7.5min, 10min, 20min, 30min, 45min, 1hour, 2hour, 3hour) to analyze. We applied different statistical measurements (mean, standard deviation, variance, median, percentile25, percentile75, interquartile range, maximum, minimum, range) to each of the collected biometric signals and the computer interaction data within the time windows selected. We then 
%compared the performance of four different machine learning 
%algorithms to be able to predict the survey responses based in the biometric and 
%computer interaction data.

%Based in the initial corpus of training data, we create a machine learning model that is able to accurately predict the responses for stress, focus, and awakeness for each of the individuals wearing the sensors.
%We are therefore able to infer the answer within the spectrum to each of these indicators without the need to ask the participants to provide an answer.
\end{comment}

@@

\subsubsection{Computer Interaction Data}
To gain a better understanding of our participants day-to-day work activities, we installed an open source computer interaction monitor~\footnote{reference omitted for double-blind review} \todo{One of the conditions to submit to Ubicomp is to NOT omit references due to anonymization purposes. So either we add the reference (which will be fine as long as we don't say we made it, or completely remove this reference)}. The monitor ran in the background on participants' computers and tracked the active windows, as well as the keyboard and mouse activity. Four of our participants opted out of this part of the study for privacy reasons (S6, S10, S12, and S13). %Therefore, we only included the computer interaction data in our analysis for our more coarse-grained explanatory model but not in our analysis on the predictive model for stress in the moment.

\subsubsection{Surveys}
\label{sec:Surveys}
Following guidelines from previous studies~\cite{Lalle16,Panwar18,Luo18} and 
following the preferrences of extensive user piloting, we sent via 
text message a survey request to each participant two times per 
work day. Pilot participants preferred the usage of text message, in part, due 
to them being accessible and noticeable anywhere in the office.
